---
title: 'Goals scored in a soccer match: fitting distributions (Ver 2)'
author: "Magdiel Ablan"
date: "07/02/2019"
always_allow_html: true
output:
  github_document:
    toc: true
    toc_depth: 2
---

## Introduction

This is part 1 of the project about goals scored in a football/soccer match. 
The objetive is to analyze the fit of 3 different distributions against observed 
data. The data are:

* Home Goals 
* Away Goals 
* Total Goals

For each of these variables 3 distributions will be fit:

* Poisson  
* Negative Binomial
* Zero-inflated Poisson


## Data preparation

First, we need to load all the required libraries:

```{r setup, echo=TRUE, message=FALSE}
# To read xlsx files:
library(readxl)

# To fit ditributions:
library(fitdistrplus)

# Add to R zip related distributions
library(VGAM)

# Legend for the plots
plot.legend <- c("Poisson", "Negative binomial","ZIP")
plot.legend2 <- c("Poisson", "Negative binomial")

# To print nicer tables
library(kableExtra)

# To know how to print tables 
outputFormat = knitr::opts_knit$get("rmarkdown.pandoc.to")

```

We now read the data. The `raw.xlsx` file contains only the columns of data 
corresponding at the goal variables. 

```{r}
raw <- read_excel("data/raw.xlsx")
head(raw)
summary(raw)
n <- length(raw$HG)
```

There are a total of `r n ` soccer games in the file. Summary stats look as
expected. 

## Home goals (HG)

`HG.sum` is table that summarises the number of home goals from all the matches
```{r}
HG.sum <-table(raw$HG)
HG.sum
HG.goals <-as.numeric(names(HG.sum))


```

The density and CDF distribution are:

```{r}
plotdist(raw$HG, histo = TRUE, discrete=TRUE)
```

We now fit the poisson distribution to this data. The only parameter of the
poisson is the mean, lambda:

```{r}
HG.pois <- fitdist(raw$HG,"pois")
HG.pois
```

Likewise, we fit the negative binomial, using the alternative parametrization
with size (target for number of successful trials) and mean (mu):

```{r}
HG.nbinom <- fitdist(raw$HG,"nbinom")
HG.nbinom
```

Finally, we fit the ZIP distribution using the functions provided in the `countreg`
package. Since this is not one of the starndard distributions provided in the
`fitdistrplus` package, we need to provide starting values for their parameters
$\pi$ and $\lambda$. These are momentum estimators based on the mean and variance
of the data

```{r}
m <- mean(raw$HG)
s <- var(raw$HG)
lambdaI <- (s + m^2)/m - 1
lambdaI
piI <- (s-m)/(s+m^2-m)
piI

```

Provided with these initial values we continue with the `fitdistr`function as
before:

```{r}
HG.zip <-fitdist(raw$HG,"zipois",start=list(lambda=lambdaI,pstr0=piI))
HG.zip

```

The maximum likelihood estimators returned by the `fitdistr` are very similar
to the previous ones. Also, as seen shortly, provide a good fit of the data so
we can safely ignore the warning.


We now compare graphically the fit of these three distributions:

```{r}
par(mfrow=c(2,2))
denscomp(list(HG.pois, HG.nbinom, HG.zip), fittype = "o" ,legendtext = plot.legend,main="HG: Histogram
         and theoretical densities")
cdfcomp(list(HG.pois, HG.nbinom,HG.zip), legendtext = plot.legend,main="HG: Empirical
         and theoretical CDFs") 
ppcomp(list(HG.pois, HG.nbinom,HG.zip), legendtext = plot.legend,main="HG: P-P plot")
qqcomp(list(HG.pois, HG.nbinom,HG.zip), legendtext = plot.legend,main="HG: Q-Q plot")
```

All the distribution seem very close to the data. However, a zoom in the first 
graph shows an improvement with the ZIP adjusted model specially for the values
of zero and one:

```{r}
denscomp(list(HG.pois, HG.nbinom,HG.zip), fittype = "o" ,legendtext = plot.legend,main="HG: Histogram
         and theoretical densities")

```

More formally, we can run goodness of fit statistics to evaluate which one has
the best fit:

```{r}
gofstat(list(HG.pois, HG.nbinom,HG.zip),fitnames=c("Poisson","Negative binomial","ZIP"))
```

These results suggest that both the negative binomial and the ZIP distribution
are good models for the data. 

We now repeat this analysis for the two other variables in the data set, away team
goals and total goals.


## Away goals (AG)

`AG.sum` is table that summarises the number of away goals from all the matches
```{r}
AG.sum <-table(raw$AG)
AG.sum
AG.goals <-as.numeric(names(AG.sum))

```

The density and CDF distribution are:

```{r}
plotdist(raw$AG, histo = TRUE, discrete=TRUE)
```

We now fit the poisson distribution to this data. The only parameter of the
poisson is the mean, lambda:

```{r}
AG.pois <- fitdist(raw$AG,"pois")
AG.pois
```

Likewise, we fit the negative binomial, using the alternative parametrization
with size (target for number of successful trials) and mean (mu):

```{r}
AG.nbinom <- fitdist(raw$AG,"nbinom")
AG.nbinom
```

Finally, we fit the ZIP distribution. But before, we need to calculate initial 
values for the parameters:

```{r}
m <- mean(raw$AG)
s <- var(raw$AG)
lambdaI <- (s + m^2)/m - 1
lambdaI
piI <- (s-m)/(s+m^2-m)
piI

```
```{r}
AG.zip <-fitdist(raw$AG,"zipois",start=list(lambda=lambdaI,pstr0=piI))
AG.zip

```

As before, only a slight modification with respect to previous values. We can 
safely ignore the warnings produced. 

We now compare graphically the fit of these distributions:

```{r}
par(mfrow=c(2,2))
denscomp(list(AG.pois, AG.nbinom,AG.zip), fittype = "o" ,legendtext = plot.legend,main="AG: Histogram
         and theoretical densities")
cdfcomp(list(AG.pois, AG.nbinom,AG.zip), legendtext = plot.legend,main="AG: Empirical
         and theoretical CDFs") 
ppcomp(list(AG.pois, AG.nbinom,AG.zip), legendtext = plot.legend,main="AG: P-P plot")
qqcomp(list(AG.pois, AG.nbinom,AG.zip), legendtext = plot.legend,main="AG: Q-Q plot")

```

Let's zoom in the first graph to see better:

```{r}
denscomp(list(AG.pois, AG.nbinom,AG.zip), fittype = "o" ,legendtext = plot.legend,main="AG: Histogram
         and theoretical densities")

```

More formally, we can run goodness of fit statistics to evaluate which one has
the best fit:

```{r}
gofstat(list(AG.pois, AG.nbinom,AG.zip),fitnames=c("Poisson","Negative binomial","ZIP"))
```

This time the ZIP distribution seem to be the best model for the data. 

## Total goals (TOT)

`TOT.sum` is table that summarises the number of away goals from all the matches
```{r}
TOT.sum <-table(raw$TOT)
TOT.sum
TOT.goals <-as.numeric(names(TOT.sum))

```

The density and CDF distribution are:

```{r}
plotdist(raw$TOT, histo = TRUE, discrete=TRUE)
```

We now fit the poisson distribution to this data. The only parameter of the
poisson is the mean, lambda:

```{r}
TOT.pois <- fitdist(raw$TOT,"pois")
TOT.pois
```

Likewise, we fit the negative binomial, using the alternative parametrization
with size (target for number of successful trials) and mean (mu):

```{r}
TOT.nbinom <- fitdist(raw$TOT,"nbinom")
TOT.nbinom
```

The size parameter is very large, specially as compared to the values obtained
for the home and away goal distribution. 

The density plot for total goals does not suggest that the ZIP distribution can
be a good model for this data. In fact, when we try to find initial values for
the parameters this is what we get:
```{r}
m <- mean(raw$TOT)
s <- var(raw$TOT)
lambdaI <- (s + m^2)/m - 1
lambdaI
piI <- (s-m)/(s+m^2-m)
piI


```

$\pi$ cannot be a negative number. Therefore, for the total goal distribution 
we only fit Poisson and negative binomial models.

We now compare graphically the fit of these two distributions:

```{r}
par(mfrow=c(2,2))
denscomp(list(TOT.pois, TOT.nbinom), fittype = "o" ,legendtext = plot.legend2,main="TOT: Histogram
         and theoretical densities")
cdfcomp(list(TOT.pois, TOT.nbinom), legendtext = plot.legend2,main="TOT: Empirical
         and theoretical CDFs") 
ppcomp(list(TOT.pois, TOT.nbinom), legendtext = plot.legend2,main="TOT: P-P plot")
qqcomp(list(TOT.pois, TOT.nbinom), legendtext = plot.legend2,main="TOT: Q-Q plot")

```

Let's zoom in the first graph to see better:

```{r}
denscomp(list(TOT.pois, TOT.nbinom), fittype = "o" ,legendtext = plot.legend2,main="TOT: Histogram
         and theoretical densities")

```

More formally, we can run goodness of fit statistics to evaluate which one has
the best fit:

```{r}
gofstat(list(TOT.pois, TOT.nbinom),fitnames=c("Poisson","Negative binomial"))
```

This shows an almost technical tie between the two. Based on the 
chi-square test, the distribution of best fit would be the negative binomial.
However, the rather large size value might be an indication that perhaps the
Poisson distribution is a better model for this variable.

## Expected goals for the Poisson distribution

This is the table for expected goals for the Poisson distribution using the data
on `r n` games and the parameters estimated in the previous sections.

```{r}
max.goals <- max(HG.goals,AG.goals,TOT.goals)

# pois.dist has the expected goal count for poisson
pois.dist <- data.frame(goals=0:max.goals,HG=numeric(max.goals+1), 
                        AG=numeric(max.goals+1),TOT = numeric(max.goals+1))

for (i in 0:(max.goals)) {
  j <- i+1
 pois.dist$HG[j] <- dpois(pois.dist$goals[j],HG.pois$estimate["lambda"])*n
 pois.dist$AG[j] <- dpois(pois.dist$goals[j],AG.pois$estimate["lambda"])*n
 pois.dist$TOT[j] <- dpois(pois.dist$goals[j],TOT.pois$estimate["lambda"])*n
}

pois.dist
```

## Expected goals for the negative binomial distribution

This is the table for expected goals for the negative binomial distribution using 
the data on `r n` games and the parameters estimated in the previous sections.

```{r}
nbinom.dist <- data.frame(goals=0:max.goals,HG=numeric(max.goals+1), 
                        AG=numeric(max.goals+1),TOT = numeric(max.goals+1))

for (i in 0:max.goals) {
  j <- i+1
  nbinom.dist$HG[j] <- dnbinom(nbinom.dist$goals[j],size = HG.nbinom$estimate["size"],
                               mu=HG.nbinom$estimate["mu"])*n
  nbinom.dist$AG[j] <- dnbinom(nbinom.dist$goals[j],size = AG.nbinom$estimate["size"],
                               mu=AG.nbinom$estimate["mu"])*n
  nbinom.dist$TOT[j] <- dnbinom(nbinom.dist$goals[j],size = TOT.nbinom$estimate["size"],
                                mu=TOT.nbinom$estimate["mu"])*n
}

nbinom.dist
```


## Expected goals for the ZIP distribution

This is the table for expected goals for the ZIP distribution using the data
on `r n` games and the parameters estimated in the previous sections.


```{r}
zip.dist <- data.frame(goals=0:max.goals,HG=numeric(max.goals+1), 
                        AG=numeric(max.goals+1))

for (i in 0:max.goals) {
  j <- i+1
  zip.dist$HG[j] <- dzipois(zip.dist$goals[j],lambda = HG.zip$estimate["lambda"],
                               pstr0=HG.zip$estimate["pstr0"])*n
  zip.dist$AG[j] <- dzipois(zip.dist$goals[j],lambda = AG.zip$estimate["lambda"],
                               pstr0=AG.zip$estimate["pstr0"])*n
  
}

zip.dist


```

## By way of conclusion

The following table summarizes the fit of the given models:

```{r}

lambdas<-round(c(HG.pois$estimate["lambda"],AG.pois$estimate["lambda"],TOT.pois$estimate["lambda"]),2)
mus <- round(c(HG.nbinom$estimate["mu"],AG.nbinom$estimate["mu"],TOT.nbinom$estimate["mu"]),2)
sizes <-round(c(HG.nbinom$estimate["size"],AG.nbinom$estimate["size"],TOT.nbinom$estimate["size"]),2)
pis <- round(c(HG.zip$estimate["pstr0"],AG.zip$estimate["pstr0"],NA),2)
Zlambdas<-round(c(HG.zip$estimate["lambda"],AG.zip$estimate["lambda"],NA),2)
best<-c("Negative binomial or ZIP","ZIP","Poisson")
sum.tab <- data.frame(lambda=lambdas,mu=mus,size=sizes,pi=pis,zlambda=Zlambdas,best=best)
row.names(sum.tab)<-c("Home","Away","Total")
kable(sum.tab,col.names=c("$\\lambda$", "$\\mu$", "n", "$\\pi$","$\\lambda$","Best model"),
      booktabs=T,escape=T)  %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "Poisson" = 1, "Negative binomial" = 2, "ZIP" = 2," "=1),align="c")
```



